---
title: "DADA2 Amplicon Pipeline — 16S Tutorial"
author: "Lab Tutorial"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{css, echo=FALSE}
body {
  font-size: 15px;
  line-height: 1.7;
}
.callout-note {
  background: #eaf4fb;
  border-left: 4px solid #3498db;
  padding: 10px 16px;
  margin: 16px 0;
  border-radius: 3px;
}
.callout-warn {
  background: #fef9e7;
  border-left: 4px solid #f39c12;
  padding: 10px 16px;
  margin: 16px 0;
  border-radius: 3px;
}
.callout-danger {
  background: #fdedec;
  border-left: 4px solid #e74c3c;
  padding: 10px 16px;
  margin: 16px 0;
  border-radius: 3px;
}
div.callout-ip34 {
  background-color: #fff4e6;   /* warm amber tint */
  border-left: 5px solid #e07b00;
  padding: 0.75em 1em;
  margin: 1em 0;
  border-radius: 4px;
}

div.callout-ip34::before {
  content: "🖥️  ip34 note";
  font-weight: bold;
  color: #e07b00;
  display: block;
  margin-bottom: 0.4em;
}
h1, h2, h3 { color: #2c3e50; }
code { color: #c0392b; }
```

---

<div class="callout-note">
**About this tutorial**  
This pipeline is adapted for paired-end Illumina amplicon data, with a focus on 16S rRNA (V5–V7 region; primers 799F/1115R). For more details on the DADA2 algorithm, refer to the [original DADA2 tutorial](https://benjjneb.github.io/dada2/tutorial.html) and the [Callahan et al. 2016 paper](https://www.nature.com/articles/nmeth.3869).
</div>

---

# Setup {#step0}
```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE)
```

## Libraries & Environment

The pipeline relies on a handful of core packages. `mgx.tools` is an in-house package that wraps several verbose utility steps (primer counting, cutadapt execution, chimera reporting, read tracking) so that the main pipeline script stays focused on the actual data processing decisions.

```{shell ip34}
# On UdeS' ip34 clusters:
ml StdEnv/2023 r/4.4.0
nice R
```


```{r setup}
library(devtools)
# install_github("trinker/pacman")
library(pacman)
# BiocManager::install("dada2")
# devtools::install_github("jorondo1/mgx.tools")
p_load(dada2, tidyverse, mgx.tools, Biostrings, ShortRead, parallel, readxl,
       update = FALSE)
```

## Configuration

All user-defined parameters are set in one place at the top of the script. **This is the only section you should need to edit** for a new project.

```{r config}
# Primers — edit for your amplicon
FWD <- "AACMGGATTAGATACCCKG"  # 799F
REV <- "AGGGTTGCGCTCGTTG"     # 1115R

ncores <- 24
```

<div class="callout-ip34">
Even though we are executing R using `nice`, which distributes core usage equitably among users, 24 is a reasonably high number of cores for ip34. Whichever cluster you are on, remember to think of others using it; don't be a corehogger!
</div>

```{r config2}
# Directory structure
# --- Edit this to set the directory parent to your raw files directory:
path_data    <- normalizePath('./data/16S') 

# --- After executing the following command, manually move your fastqs there: 
path_raw     <- file.path(path_data, '0_raw')

# --- Create additional folders for the rest of the pipeline
path_raw     <- file.path(path_data, '1_filtN')
path_cut     <- file.path(path_data, '2_cutadapt')
path_out     <- file.path(path_data, '4_out')
path_summary <- file.path(path_data, 'X_DADA2_Summary')

# This is the path on ip34/Mammoth.
# Edit this to point to your install.
path_cutadapt <- '/cvmfs/soft.mugqic/CentOS6/software/cutadapt/cutadapt-2.10/bin/cutadapt'

# Create output directories if needed
if (!dir.exists(path_cut))     dir.create(path_cut)
if (!dir.exists(path_out))     dir.create(path_out)
if (!dir.exists(path_summary)) dir.create(path_summary)
```

<div class="callout-note">
**Directory numbering convention:** Directories are prefixed with integers (`0_raw`, `1_filtN`, `2_cutadapt`, `3_filtered`, `4_out`) to match the pipeline steps and keep them ordered in the file system. `X_DADA2_Summary` collects all diagnostic plots and reports.
</div>

## Input Files

DADA2 expects one pair of gzipped FASTQ files per sample, named consistently. Here we assume the pattern `SampleName_R1.fastq.gz` / `SampleName_R2.fastq.gz`.

```{r input}
fnFs <- sort(list.files(path_raw, pattern = "_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path_raw, pattern = "_R2.fastq.gz", full.names = TRUE))

# Extract sample names by stripping the read-direction suffix
(sample.names <- gsub("_R[12].*$", "", basename(fnFs)))

# Save sample list for reference
write_delim(data.frame(sample.names), file.path(path_summary, 'sample_names.tsv'))
```

**Always visually confirm `sample.names`** before proceeding. Incorrect naming will propagate through the entire pipeline.

---

# N-filtering {#step1}

Before primer removal, we remove reads containing ambiguous bases (Ns) and low-complexity reads. This is a lightweight pre-filter; it does **not** apply quality thresholds.

```{r nfilt}
fnFs.filtN <- file.path(path_data, "1_filtN", basename(fnFs))
fnRs.filtN <- file.path(path_data, "1_filtN", basename(fnRs))

out.N <- filterAndTrim(
  fnFs, fnFs.filtN,
  fnRs, fnRs.filtN,
  maxN        = 0,         # reject any read containing an N
  rm.lowcomplex = TRUE,    # remove low-complexity reads (see note below)
  verbose     = TRUE,
  multithread = ncores
)

names(fnFs.filtN) <- sample.names
names(fnRs.filtN) <- sample.names
rownames(out.N)   <- sample.names

head(out.N)
```

<div class="callout-note">
**`rm.lowcomplex = TRUE`:** Low-complexity reads (e.g. long homopolymer stretches) can cause issues downstream. This option was added following a known DADA2 issue with certain sequencing runs — see [GitHub #2045](https://github.com/benjjneb/dada2/issues/2045#issuecomment-2452299127).
</div>

---

# Primer removal (Cutadapt) {#step2}

Primers must be removed before DADA2 error modelling, because primer sequences are not biological signal and will confound the error model. We use [Cutadapt](https://cutadapt.readthedocs.io/).

## Check primer occurrence

This step checks how often each primer (in all four orientations) appears in the raw reads.

```{r primer-check-pre}
mgx.tools::primer_occurence(fnFs.filtN, fnRs.filtN, FWD, REV, ncores = ncores)
```

You should see the forward primer predominantly in R1 (`FWD.ForwardReads`) and the reverse primer predominantly in R2 (`REV.ReverseReads`). Significant presence of reverse-complements in both reads is expected for shorter amplicons where read-through occurs.

## Run Cutadapt

```{r cutadapt}
fnFs.cut <- file.path(path_cut, basename(fnFs))
fnRs.cut <- file.path(path_cut, basename(fnRs))

# Each sample is processed independently (single-core per sample, parallelised across samples)
mclapply(
  seq_along(fnFs),
  mgx.tools::run_cutadapt,
  cutadapt_path = path_cutadapt,
  # Remove FWD primer and reverse-complement of REV from R1
  R1.flags = paste("-g", FWD, "-a", dada2:::rc(REV)),
  # Remove REV primer and reverse-complement of FWD from R2
  R2.flags = paste("-G", REV, "-A", dada2:::rc(FWD)),
  mc.cores = ncores
)
```

<div class="callout-note">
**Adapter trimming flags:**  
- `-g` / `-G`: trim adapter from the **5' end** (primers)  
- `-a` / `-A`: trim adapter from the **3' end** (reverse-complements of the opposite primer, present when the amplicon is shorter than the read)  
- `-n 2`: allow up to 2 rounds of adapter trimming per read (ensures both adapters are removed when both are present)  
- The `-m 21 -M 300` length bounds (set internally in `run_cutadapt`) discard reads that are too short to be biological after trimming — see [GitHub #2045](https://github.com/benjjneb/dada2/issues/2045#issuecomment-2449416862)
</div>

## Verify primer removal

```{r primer-check-post}
mgx.tools::primer_occurence(fnFs.cut, fnRs.cut, FWD, REV, ncores = ncores)
```

All primer counts should be zero (or very close to zero). If not, check your primer sequences and orientations before continuing.

Generate a list of primer-trimmed files:

```{r cutadapt-files}
cutFs <- sort(list.files(path_cut, pattern = "_R1.fastq.gz", full.names = TRUE))
cutRs <- sort(list.files(path_cut, pattern = "_R2.fastq.gz", full.names = TRUE))
```

---

# Quality filtering {#step3}

This is where quality-based decisions are made. The goal is to balance retaining enough reads while discarding low-quality sequence that would inflate error rates.

## Inspect quality profiles (before filtering)

Quality profiles should be inspected **before** setting truncation parameters. These plots show per-position quality scores across randomly selected samples and across all samples (aggregated).

```{r qplot-raw}
plot_list <- mgx.tools::gen_qplots(Fs = cutFs, Rs = cutRs, nsam = 4)
mgx.tools::save_qplots(plot_list = plot_list,
            out_dir   = path_summary,
            step_id   = 'raw')
```

<div class="callout-warn">
**Setting `truncLen`:** This is the most dataset-specific parameter in the pipeline. Quality typically degrades near the 3' end of reads. Set truncation lengths where quality drops consistently below Q25–30. For 16S V5-V7 with 250 bp paired-end reads, values around `truncLen = c(190, 140)` are common, but **always verify with your own quality plots**.  

The truncated reads must still overlap sufficiently for merging (Step 5). A minimum overlap of 12 bp is used here, so ensure `truncLen[1] + truncLen[2] - amplicon_length ≥ 12`.
</div>

## Filter and trim

```{r filter}
filtFs <- file.path(path_data, "3_filtered", basename(cutFs))
filtRs <- file.path(path_data, "3_filtered", basename(cutRs))

names(filtFs) <- sample.names
names(filtRs) <- sample.names

out <- filterAndTrim(
  cutFs, filtFs, cutRs, filtRs,
  maxEE    = c(2, 2),       # max expected errors per read; increase to c(4,4) for less stringent filtering
  truncLen = c(190, 140),   # ADJUST based on your quality profiles
  minLen   = 50,            # discard reads shorter than 50 bp (e.g. after trimming)
  rm.phix  = TRUE,          # remove PhiX control reads
  compress = TRUE,
  multithread = ncores
)

rownames(out) <- sample.names
```

<div class="callout-note">
**`maxEE` (Maximum Expected Errors):** This filter is based on the sum of error probabilities derived from quality scores, as described in [Edgar & Flyvbjerg 2015](https://doi.org/10.1093/bioinformatics/btv401). It is generally more informative than a simple quality score cutoff. Values of `c(2, 2)` are standard; `c(4, 4)` can help retain more reads from lower-quality runs, at a modest cost in accuracy.
</div>

## Inspect quality profiles (after filtering)

```{r qplot-clean}
plot_list <- gen_qplots(Fs = filtFs, Rs = filtRs, nsam = 4)
save_qplots(plot_list = plot_list,
            out_dir   = path_summary,
            step_id   = 'clean')
```

## Handle dropped samples

Stringent length filtering can cause low-input samples (e.g. negative controls, failed extractions) to produce empty output files. DADA2 will error if it tries to process a non-existent file.

```{r dropped}
# Identify files that survived filtering; warn about any that didn't
filtFs_survived <- mgx.tools::dropped_samples(filtFs)
filtRs_survived <- mgx.tools::dropped_samples(filtRs)
```

---

# Step 4 — Error model & denoising {#step4}

The DADA2 error model is the core of the pipeline. It learns the rate at which each nucleotide substitution occurs as a function of quality score, using a subset of the data. This model is then used to distinguish true biological sequences from sequencing errors.

## Learn error rates

```{r errors}
errF <- learnErrors(filtFs_survived, multithread = ncores)
errR <- learnErrors(filtRs_survived, multithread = ncores)

# Visualise the fitted error model — this is an important QC step
err_plot <- plotErrors(errF, nominalQ = TRUE)

ggsave(filename = file.path(path_summary, "error_plot.pdf"),
       plot = err_plot, bg = 'white',
       width = 2000, height = 2000, units = 'px', dpi = 180)
```

<div class="callout-note">
**Interpreting the error plot:** Each panel shows observed error rates (points) and the fitted model (line) for one substitution type, as a function of quality score. The fit should follow the points reasonably well — rates should decrease as quality increases. If the model looks poorly fitted (flat lines, poor tracking of points), consider increasing the data used for learning errors or checking for quality issues upstream.
</div>

## Denoise (sample inference)

DADA2 uses the error model to partition reads into true sequence variants and noise.

```{r denoise}
dadaFs <- dada(filtFs_survived, err = errF, pool = 'pseudo', multithread = ncores)
dadaRs <- dada(filtRs_survived, err = errR, pool = 'pseudo', multithread = ncores)
```

<div class="callout-note">
**`pool = 'pseudo'`:** Pooling allows DADA2 to detect rare ASVs that might be missed when samples are processed independently. Full pooling (`pool = TRUE`) is the most sensitive but memory-intensive; pseudo-pooling is a good compromise that uses a two-pass heuristic to approximate pooling without loading all data simultaneously. For most datasets, pseudo-pooling offers a meaningful gain in sensitivity over `pool = FALSE` with acceptable computational cost.
</div>

---

# Step 5 — Sequence table {#step5}

## Merge paired reads

Forward and reverse reads are merged at their overlapping region. Merging fails (reads are discarded) if the overlapping bases do not match sufficiently.

```{r merge}
mergers_pooled <- mergePairs(
  dadaFs, filtFs_survived,
  dadaRs, filtRs_survived,
  minOverlap = 12    # minimum overlap in bp; increase if you have long amplicons
)
```

## Build sequence table

```{r seqtab}
seqtab <- makeSequenceTable(mergers_pooled)
table(nchar(getSequences(seqtab)))  # inspect the ASV length distribution
```

## Filter by expected amplicon length

The length distribution should show a clear peak at the expected amplicon length. Sequences outside this range are likely artefacts (non-specific amplification, chimeras not yet removed). Adjust the range to match your amplicon.

```{r length-filter}
# Keep only ASVs within the expected length range for 16S V5-V7
seqtab2 <- seqtab[, nchar(colnames(seqtab)) %in% 290:310]
```

## Remove chimeras

Chimeric sequences are PCR artefacts formed by the spurious joining of two different template molecules. They are identified and removed based on whether a sequence can be reconstructed as a bimera of two more abundant sequences.

```{r chimeras}
seqtab.nochim <- removeBimeraDenovo(
  seqtab2, method = "consensus", multithread = ncores, verbose = TRUE
)

# Report chimera statistics
sink(file.path(path_summary, "chimera_report.txt"), append = FALSE, split = TRUE)
mgx.tools::chimera_report(seqtab, seqtab.nochim)
sink()

table(nchar(getSequences(seqtab.nochim)))  # length distribution after chimera removal
```

<div class="callout-warn">
**Expected chimera rates:** In a typical 16S run, 5–25% of ASVs may be chimeric, but they usually represent a small fraction of total reads (<5%). If chimera removal eliminates a large proportion of reads (>20%), this may indicate issues with primer removal, quality filtering, or the PCR protocol itself.
</div>

## Pre-filter rare ASVs

Very rare ASVs (present in only 1–2 reads across the entire dataset) are most likely artefacts, and retaining them substantially increases the computational cost of taxonomic assignment. This is a conservative pre-filter; more thorough rarefaction-based filtering is typically applied during the downstream analysis.

```{r rare-asvs}
# Visualise the relationship between minimum count threshold and ASV retention
mgx.tools::minimum_ASV_count(seqtab.nochim) %>%
  ggsave(filename = file.path(path_summary, "rare_ASVs.pdf"),
         plot = ., bg = 'white', width = 2000, height = 2000, units = 'px', dpi = 180)

# Drop ASVs with fewer than 2 total reads across all samples
seqtab.filt <- mgx.tools::drop_rare_asvs(seqtab.nochim, at_least_n = 2)
dim(seqtab.filt)
```

## Save output

```{r save-seqtab}
write_rds(seqtab.filt, file.path(path_out, 'seqtab.RDS'), compress = 'gz')
```

---

# Step 6 — Read tracking {#step6}

A summary of read counts at each pipeline step is an essential QC tool. Large losses at a specific step point to specific problems.

```{r tracking}
track_change <- mgx.tools::track_dada(
  out.N        = out.N,
  out          = out,
  sample.names = sample.names,
  dadaFs       = dadaFs,
  mergers      = mergers_pooled,
  seqtab.nochim = seqtab.nochim
)

library(patchwork)
mgx.tools::plot_track_change(track_change) %>%
  ggsave(filename = file.path(path_summary, "track_changes.pdf"),
         plot = ., bg = 'white',
         width = 1600, height = 2200, units = 'px', dpi = 180)
```

**What to look for:**

| Step | Acceptable loss | Warning sign |
|---|---|---|
| N-filtering | < 5% | > 10%: quality/sequencing issue |
| Cutadapt | < 10% | > 20%: wrong primer sequences |
| filterAndTrim | 20–50% | > 70%: overly stringent truncation or quality issue |
| Denoising | < 5% | > 20%: poor error model fit |
| Merging | < 20% | > 50%: truncLen too aggressive; insufficient overlap |
| Chimera removal | < 10% of reads | > 30%: PCR or primer issues |

---

# Step 7 — Taxonomic assignment {#step7}

<div class="callout-danger">
**⚠️ Read before running taxonomy**

DADA2's native `assignTaxonomy()` uses a naïve Bayes classifier trained on a reference database. It works reasonably well for well-characterised markers like 16S, but has important limitations:

- **It is not recommended for non-16S amplicons.** Classification accuracy drops substantially for ITS, *trnL*, and especially fungal/AMF markers, where reference databases are less comprehensive and the naive Bayes approach struggles with short or variable-length amplicons.
- **DECIPHER's `IdTaxa` is generally preferred**, especially for anything other than 16S. It is a machine learning approach with better calibrated confidence scores. See the [DECIPHER tutorial](http://www2.decc.nl/bioinformatics/DECIPHER/) and [Murali et al. 2018](https://doi.org/10.1186/s40168-018-0521-5).
- For 16S, `assignTaxonomy()` is acceptable, but still consider using `IdTaxa` for better confidence estimates.
</div>

## Assign taxonomy (SILVA — 16S)

```{r taxonomy}
# Load seqtab if continuing from a saved session
seqtab_16S <- read_rds(file.path(path_out, 'seqtab.RDS'))

silva_train  <- "path/to/silva_nr99_v138.1_train_set.fa.gz"
silva_species <- "path/to/silva_species_assignment_v138.1.fa.gz"

# Genus-level classification
tax_Genus <- assignTaxonomy(
  seqtab_16S, silva_train,
  tryRC = TRUE, multithread = TRUE, verbose = TRUE
)

# Species-level exact matching
tax_Species <- addSpecies(
  tax_Genus, silva_species, tryRC = TRUE
)
```

<div class="callout-note">
**`tryRC = TRUE`:** Tells DADA2 to also try classifying the reverse complement of each ASV. This is important when your amplicon can be sequenced in either orientation — common with certain primer pairs or when reads are not consistently oriented.
</div>

## Build phyloseq object

```{r phyloseq}
# Clean up sample names if needed (e.g. remove amplicon suffix)
sample_names_clean <- gsub("_16S", "", rownames(seqtab_16S))
rownames(seqtab_16S) <- sample_names_clean

ps <- phyloseq(
  tax_table(tax_Species),
  otu_table(seqtab_16S, taxa_are_rows = FALSE)
)

write_rds(ps, file.path(path_out, 'ps.rds'))
```

---

# Step 8 — Classification rate {#step8}

A useful sanity check: what proportion of ASVs (and reads) could be assigned at each taxonomic rank?

```{r classrate}
ps <- readRDS(file.path(path_out, 'ps.rds'))

melted <- ps %>%
  psmelt() %>%
  filter(Abundance > 0) %>%
  group_by(Sample) %>%
  mutate(relAb = Abundance / sum(Abundance))

ranks <- c('Family', 'Genus', 'Species')

classification <- map(ranks, function(rank) {
  melted %>%
    filter(relAb > 0) %>%
    select(Sample, !!sym(rank), relAb) %>%
    mutate(classified = case_when(is.na(!!sym(rank)) ~ 0, TRUE ~ 1)) %>%
    summarise(
      asv_prop   = sum(classified) / n(),          # proportion of ASVs classified
      relAb_prop = sum(classified * relAb)         # abundance-weighted proportion
    ) %>%
    pivot_longer(cols = c('relAb_prop', 'asv_prop'), names_to = 'proportion_type') %>%
    mutate(taxRank = factor(rank, levels = ranks))
}) %>% list_rbind()

classification %>%
  filter(proportion_type == 'asv_prop') %>%
  mutate(proportion_type = case_when(
    proportion_type == 'asv_prop'   ~ 'Proportion of ASVs',
    proportion_type == 'relAb_prop' ~ 'Proportion of reads'
  )) %>%
  ggplot(aes(y = value, x = taxRank, fill = taxRank)) +
  geom_line(aes(group = Sample), linewidth = 1, colour = 'grey') +
  geom_boxplot() +
  scale_colour_brewer(palette = 'Set2') +
  theme_minimal() +
  labs(y = 'Proportion of classified ASVs') +
  theme(
    axis.ticks.x    = element_blank(),
    axis.title.x    = element_blank(),
    legend.position = 'none'
  )

ggsave(file.path(path_summary, '16S_classrate.pdf'),
       bg = 'white', width = 1600, height = 1600,
       units = 'px', dpi = 260)
```

<div class="callout-note">
**Interpreting classification rates:**  
- Family-level classification above ~80–90% is typical for well-characterised ecosystems with 16S.  
- Genus-level classification drops depending on the database version and the microbial community.  
- Species-level (`addSpecies`) relies on exact sequence matching and is typically low (<30%) — this is expected behaviour, not a problem.  
- Consistently low family-level classification (< 60%) may indicate: off-target amplification, a poorly suited reference database, or a truly novel community (e.g. environmental samples from poorly characterised habitats).
</div>

---

# Appendix — Adapting to other amplicons {#appendix}

This pipeline applies to other amplicons with minimal changes. Key considerations:

| Amplicon | Primer change | `truncLen` | Length filter | Taxonomy |
|---|---|---|---|---|
| ITS1/ITS2 | Yes | Often none (variable length — set `truncLen = 0`) | Wider range | UNITE; prefer `IdTaxa` |
| *trnL* (P6) | Yes | Often none | Narrow range | Custom DB |
| *chiITS2* (oomycetes) | Yes | Yes | Yes | Custom DB; `IdTaxa` strongly preferred |
| AMF (AML2/WANDA) | Yes | Yes | Yes | MaarjAM DB; `IdTaxa` strongly preferred |

<div class="callout-danger">
**For ITS and other variable-length amplicons:** Do **not** set `truncLen` — truncating ITS reads at a fixed length is incorrect because the ITS region has highly variable length and truncation will create spurious length artefacts. Instead, rely entirely on `maxEE` for quality filtering, and use `minLen` to discard very short reads.
</div>

---

*For questions or issues with this pipeline, refer to the [DADA2 GitHub issues page](https://github.com/benjjneb/dada2/issues) for known bugs, or the [DADA2 forum](https://forum.qiime2.org/) for community support.*
